A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-20 paper/paper.tex
Add definition for the recursor

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
      [ [[Void]] ]_i &= \emptyset \\
      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text { and } [[a<=>b]] \}\\
    \end{split}
  \end{equation*}
  \caption{Logical relation for \lang}
  \label{fig:logrel}
\end{figure}

In this section, we show the definition of the logical relation as a
partial function in Figure~\ref{fig:logrel}. The logical relation can
be viewed as a partial function that takes a closed type,
a natural number $[[i]]$ representing the universe level, and
returns a set of closed terms that \emph{semantically} inhabit the
closed type, though we do not strictly enforce the closedness of
either the type or the term.

>>>>

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
      [ [[Void]] ]_i &= \emptyset \\
      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
    \end{split}
  \end{equation*}
  \caption{Specification of the logical relation for \lang}
  \label{fig:logrel}
\end{figure}

In this section, we present the definition of the logical relation as a
partial function.
Figure~\ref{fig:logrel} shows the specification
of the logical relation that we are about to define. The specification
itself does not immediately induce a definition as it is unclear
whether there exists a smallest partial function that satisfies the
specification.
Without such as proof, we would not be able to induct on the graph of
the function or use any inversion principle.

The construction of a function that meets the specification in
Figure~\ref{fig:logrel} can be broken down into two steps. To build an
infinite hierarchy of universes, we first define a recursor that constructs
the interpretation of the universe level $[[n]]$ given a parameterized
interpretation function for universe levels that are strictly less
than $[[n]]$. The well-definedness of this inductive relation requires
some extra justification as it is not strictly positive. Instead, we
must manually prove the monotonicity of the underlying functor.
Next, we use the well-foundedness of the less than relation to tie the
knot and define the full interpretation function that meets the
specification.

For concision, we first fix the following parameters:
\begin{equation*}
  \begin{split}
    i &:: \mathit{Nat} \\
    I &:: \mathit{Nat} \rightarrow \mathit{Term} \times \mathcal{P}(\mathit{Term})
  \end{split}
\end{equation*}

Now, consider the following function $F$:
\begin{equation*}
  \begin{split}
    F &:: \mathit{Term} \times \mathcal{P}(\mathit{Term}) \rightarrow \mathit{Term} \times \mathcal{P}(\mathit{Term})\\
    F(X) &= F_V  \cup F_B \cup F_I \cup F_\Pi(X) \cup F_U \cup F_S(X) \\
    F_V &= \{ ([[Void]], \emptyset) \} \\
    F_B &= \{ ([[Bool]], \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a
          =>+ false]] \})  \} \\
    F_I &= \{ ([[a ~ b : A]], \{ [[p]] \mid [[p =>+ refl]] \text{ and
          } [[a<=>b]] \})  \} \\
          % This definition is wrong because it'd include all pi types
    F_{\Pi}(X) &= ([[Pi A B]], \{ [[b]] \mid \exists A\text{ and }S, (A,S) \in X,
                   \forall [[a]], \text{ if } [[a]] \in S \text{, then
                 } \\
    F_U &= \{([[Set j]] , \{ [[A]] \mid [[A]] \in
          dom(I_j) \}) \mid j < i \} \\
    F_S(X) &= ([[A]], \{ [[a]] \mid \exists ([[B]], S) \in X, [[a]]
          \in S \text{ and } [[A => B]] \})  \\
  \end{split}
\end{equation*}

%  The logical relation can
% be viewed as a partial function that takes a closed type,
% a natural number $[[i]]$ representing the universe level, and
% returns a set of closed terms that \emph{semantically} inhabit the
% closed type, though we do not strictly enforce the closedness of
% either the type or the term.

% The claim that $[ - ]_=$ is a partial function is not immediately
% obvious due to the last equation in its definition. The last equation
% ensures that the if some type $[[A]]$ is in the domain of the
% interpretation function, then any type $[[B]]$ that steps into $[[A]]$
% has the same interpretation as $[[A]]$. For example,
% since $[[if true Bool Void]]$ steps into $[[Bool]]$, we can say that
% $[[if true Bool Void]]$ is semantically inhabited by terms that
% reduces down to either $[[true]]$ or $[[false]]$ through the
% $[[Bool]]$ case of the interpretation function.

% However, due to the non-deterministic nature of parallel reduction,
% given some terms $[[A]]$, $[[B0]]$, $[[B1]]$ such that $[[A => B0]]$
% and $[[A => B1]]$, the meaning of $[[A]]$ could either be $[ [[B0]] ]_i$
% or $[ [[B1]] ]_i$. In fact, the definedness of $[ [[B0]] ]_i$ and $[
% [[B1]] ]_i$
% should not be taken as granted either. It is unclear for now whether
% the definedness of one implies the other.

% Fortunately,
